---
title: "Introduction"
output: html_document
---
```{r echo=FALSE}
knitr::opts_chunk$set(eval=evaluate, cache=cache.me)
```

This R-script and documentation was created to calculate the Phytoplankton Index of Biotic Integrity (PIBI) for the Chesapeake Bay and its tidal tributaries developed by [@lacouture_phytoplankton_2006]. The following text documents the process used to prepare and calculate the PIBI. Data was acquired from the Chesapeake Bay Program's (CBP) CEDR database.

### Intended Audience

I developed this document for the Chesapeake Bay Program and individuals/groups interested in assessing the Chesapeake Bay condition with phytoplankton. The document describes the how to use the statistical coding language, [@r_core_team_r:_2017], to re-create the PIBI. My hope is that this document will allow others to assess Chesapeake Bay conditions based on the indices developed by [@lacouture_phytoplankton_2006]. 

This document provides the R code necessary to import, prepare, and calculated the indices developed by [@lacouture_phytoplankton_2006]. __R Markdown__ is used to integrate text and code. The intention is to provide a more readable document, with better descriptions of how and why an activity was performed, than a standard R script with a few commented lines. Although my hope is that readers without any R background could comprehend the R code in this document with a little bit of patience, this document is really intended for readers with at least a basic understanding of R programming. 

Additionally, I am a big fan of the [tidyverse packages](https://www.tidyverse.org/packages/) developed by the [RStudio](https://www.rstudio.com/) team and use these packages frequently throughout the document. The [tidyverse](https://www.tidyverse.org/packages/) is essentially an ecosystem of packages that work well together and often make it easier to deal with data in R. For R users who do not use packages from the [tidyverse](https://www.tidyverse.org/packages/), I think many of the functions will be intuitive or easy to quickly search for online. However, I often find that the pipe operator, `%>%`, from the [magrittr package](http://magrittr.tidyverse.org/) is confusing to those unfamiliar with the [tidyverse](https://www.tidyverse.org/packages/), and right fully so. It takes a little while to wrap your head around the pipe operator but once you do I think you will find its use makes R code more legible. In essence base R works from the inside-out, while the pipe operator presents the code in a linear fashion. For example, imagine you have a character vector `x` and you want to trim leading/trailing white space, then keep only unique strings, and then make all characters lowercase. In base R, the code would look this: `tolower(unique(trimws(x)))`. Again, base R works from the inside-out, so first `trimws()` is used to remove leading/trailing white space, then `unique()` is used to remove duplicate strings, and finally `tolower()` is used to convert all characters to lowercase. Using the pipe operator, the code would look like this: `x %>% trimws() %>%  unique() %>% tolower()`. The pipe operator presents the functions in the order you intend them to be performed; therefore, the code should be easier to read and more intuitive, which should in turn reduce errors. For more information on the pipe operator please refer to [http://magrittr.tidyverse.org/](http://magrittr.tidyverse.org/).

### Download and Edit this Document

This document can be downloaded and edited from my [GitHub account](https://github.com/zsmith27/phyto). As stated in the previous section, this document was created using __R Markdown__, so that the code could be integrated with descriptive text. The document is broken up into multiple children documents in the "notebook" folder, which are all compiled together using the "pipi_notebook.Rmd" parent document. For small corrections, the easiest way to inform me will be to [open a new issue on the GitHub respository for this document](https://github.com/zsmith27/phyto/issues). However, my hope is that there will be others interested in becoming more involved, of whom I can add as collaborators to the Git Hub repository. [RStudio has made it very easy to connect R projects to GitHub](https://support.rstudio.com/hc/en-us/articles/200532077-Version-Control-with-Git-and-SVN). Using these tools, it can be very easy to "push" and "pull" edits, while maintaining a record of the documents history. If you are interested in collaborating on this document, please contact me, [Zachary M. Smith](https://www.potomacriver.org/about-us/staff/).


### Getting Started

Load the necessary packages into the environment [@dowle_data.table:_2017; @wickham_httr:_2017; @grolemund_dates_2011; @wickham_stringr:_2017; @wickham_readxl:_2017; @lang_rcurl:_2016; @wickham_dplyr:_2017; @wickham_tidyr:_2017; @henry_purrr:_2017] __ADD jsonlite and rprojroot__. 
```{r load_packages, message=FALSE}
library(tidyverse)
library(lubridate)
library(readxl)
library(httr)
library(RCurl)
library(data.table)
library(jsonlite)
library(rprojroot)
```

__mmir__ is a costume package that I am developing to calculate taxonomic community metrics. The development version of __mmir__ can be downloaded from [my GitHub account](https://github.com/zsmith27/mmir) using `devtools::install_github(repo = "zsmith27/mmir")`. After the installation __mmir__ must be loaded into the environment using `library()`.
```{r}
library(mmir)
```

`sessionInfo()` is used to provide the information necessary to replicate this document. As packages are updated, especially __mmir__, functions may become defunct or a function may not respond in a similar fashion to previous versions of the package. Without the information provided by `sessionInfo()` it can be extremely difficult, if not impossible to replicate the results in this report.
```{r}
sessionInfo()
```

Data for this document are located in various files. The object `project.dir` represents the phytoplankton project directory and will be used as the root of the file path throughout the document.
```{r}
project.dir <- rprojroot::find_rstudio_root_file()
```

### Custom Functions

The `clean_string()` function modifies character objects by removing leading/trailing white space, converting all characters to lower case, and replacing all spaces (" ") with an underscore ("_"). These specifications eliminate common typos and standardize character objects.
```{r}
clean_string <- function(x) {
  x %>% 
    stringr::str_trim() %>% 
    tolower() %>% 
    stringr::str_replace_all("\\s+", " ") %>% 
    stringr::str_replace_all(" ", "_") %>%  
    if_else(. == "", as.character(NA), .)
}
```

The `clean_up()` function is a wrapper for `clean_string()`. This function applies `clean_string()` to the column headers and character columns of data frames. Again, this process standardizes character values and makes it easier work with the data.
```{r}
clean_up <- function(x) {
  x %>% 
    rename_all(clean_string) %>% 
    mutate_if(is.character, funs(clean_string))%>% 
    distinct()
}
```

