---
title: "Prepare Carbon Assignments"
output: html_notebook
---

```{r echo=FALSE}
knitr::opts_chunk$set(eval=evaluate, cache=TRUE)
```

A large but incomplete table of taxa size and carbon values is imported from the project data directory.
```{r}
carbon.df <- readxl::read_excel(file.path(project.dir, "data/carbon/carbon_list_2014.xlsx"),
                                sheet = "carbon_list_2014") %>% 
  clean_up()
  
```

Some of the `size` categories are standardized to make it possible to accurately join `carbon.df` and `bay.df`. "unidentified" is converted to `NA`. Both `carbon.df` and `bay.df` have inconsistent representation of "species" (i.e. "species", "species_", "species_#", "sp.", and "sp#"): therefore, "species" was standardized to "sp#".
```{r}
carbon.df <- carbon.df %>% 
  mutate(size = case_when(
    size %in% "unidentified" ~ as.character(NA),
    str_detect(size, "species|sp.") ~ str_replace_all(size, "sp.#|species_#|species_|species", "sp#"),
    TRUE ~size
  )) %>% 
  mutate(picoplankton = if_else(
    diameter <= 20 &
      height <= 20 &
      width <= 20 &
      width <= 20 &
      height_2 <= 20 &
      width_2 <= 20,
    TRUE,
    FALSE
  )) %>% 
  select(latinname, size, old_carbon) %>%
  distinct()
```

Check for taxa with duplicate carbon values. If a taxon have two or more carbon values, then when `carbon.df` is joined with `bay.df` it will duplicate the taxon's reporting value.
```{r}
carbon.dups <- carbon.df %>% 
  group_by(latinname, size) %>% 
  mutate(count = n()) %>% 
  ungroup() %>% 
  filter(count > 1)
```

Taxa with duplicate carbon values: `r unique(carbon.dups$latinname)`. One of the "chaetoceros_wighami" `old_carbon` values is `NA`, and therefore, is excluded. The selected `old_carbon` values for "biddulphia" and "gymnodinium" appear to be the values used by [Lacouture et al. (2006)](https://www.potomacriver.org/focus-areas/aquatic-life/phytoplankton-index-of-biotic-integrity-for-chesapeake-bay/).
```{r}
carbon.df <- carbon.df %>% 
  filter(!(latinname == "chaetoceros_wighami" & is.na(old_carbon)),
         !(latinname == "biddulphia" & old_carbon == 7899.50), # %>%  #,
         !(latinname == "gymnodinium" & old_carbon == 848))
#  mutate(size = case_when(
#    is.na(size) & latinname == "gymnodinium" & old_carbon == 848 ~ "msu",
#    is.na(size) & latinname == "gymnodinium" & old_carbon == 43.900 ~ "odu",
#    TRUE ~ size
#    ))
         
```

Convert string values "not_applicable" and "not_specified" in the `size` column to `NA`. Also, some "gymnodinium" had a reported size of "cell-_small", which had no equivalent size category in `carbon.df`; therefore, "cell-_small" is replaced with `NA`. Finally, as was done above for `carbon.df`, strings representing "species" were standardized to "sp#".
```{r}
bay.df <- bay.df %>% 
  mutate(size = case_when(
    size %in% c("not_applicable", "not_specified") ~ as.character(NA),
    size == "cell-_small" & latinname == "gymnodinium" ~ as.character(NA),
    str_detect(size, "species|sp.") ~ str_replace_all(size, "sp.#|species_#|species_|species", "sp#"),
    TRUE ~ size
  ))# %>% 
#  mutate(size = case_when(
#        is.na(size) & latinname == "gymnodinium" & str_detect(source, "msu") ~ "msu",
#    is.na(size) & latinname == "gymnodinium" & str_detect(source, "odu") ~ "odu",
#    TRUE ~ size
#  ))
```

Identify all taxa (`latinname`) in `bay.df` that do not have a size equivalent in `carbon.df`.
```{r}
no.match.df <- anti_join(bay.df, carbon.df, by = c("latinname", "size")) %>% 
  select(latinname, size) %>% 
  distinct()
```

Attempt to find partial matches for taxa in `no.match.df`. Each row in `no.match.df` represents a unique taxon and size combination. `lapply` loops through each row, extracts the unique `size` string, splits the string into smaller pieces, and attempts to find partial matches with associated `size` strings in `carbon.df`. The string in `carbon.df` with the most partial matches will be used to replace the `size` string in `bay.df`.
```{r}
partial.match.df <- lapply(1:nrow(no.match.df), function(row.i) {
  sub.df <- no.match.df %>% 
    slice(row.i)
  
  size.vec <- str_split(sub.df$size, "_") %>% unlist()
  
  getmode <- function(v) {
    uniqv <- unique(v)
    uniqv[which.max(tabulate(match(v, uniqv)))]
  }
  
  
  carbon.sub <- carbon.df %>% 
    filter(latinname == sub.df$latinname) %>% 
    mutate(row_max_matches = sapply(size.vec, function(size.i) grep(size.i, size)) %>% 
             unlist() %>% 
             getmode()) %>% 
    slice(unique(row_max_matches)) %>% 
    select(-row_max_matches, -old_carbon) %>% 
    rename(new_size = size) %>% 
    mutate(size = sub.df$size)
  
  return(carbon.sub)
}) %>% 
  bind_rows()
```

Merge `bay.df` with `partial.match.df` and if a `new_size` is present from `partial.match.df`, then replace the `size` string with the `new_size` string. By updating the `size` strings it is possible to obtain more matches when merging `bay.df` and `carbon.df`.
```{r}
bay.df <- left_join(bay.df, partial.match.df, by = c("latinname", "size")) %>% 
  mutate(reported_size = size,
         size = if_else(!is.na(new_size), new_size, size))
```

Merge `bay.df` with `carbon.df` and calculate biomass. The phytoplankton counts are reported in picoliters. Biomass is converted to microliters.
```{r}
bay.df <- left_join(bay.df, carbon.df, by = c("latinname", "size")) %>% 
  mutate(biomass = reportingvalue * old_carbon / 10^6)
```





