---
title: "R Notebook"
output: html_notebook
---

```{r echo=FALSE}
knitr::opts_chunk$set(eval=evaluate, cache=cache.me)
```

### Metric Disagreement


This section compares the values found in this document to the values found by Jacqueline M. Johnson. The previous section was good for identifying issues related to scoring the metrics. This section identifies issues related to the metric values. The discrepancies found here most likely do not have to deal with the formula used to calculate the metrics but probably stem from differences in how the taxonomic counts are prepared.

Values from this document (`ratings.df`) are merged with Jacqueline M. Johnson's values (`old.df`).  Season (`season`) and salinity zone (`salzone`) are combined (`unite()`) to represent phytoplankton index names. Metrics (`metric`), indices (`index`), and data sources (`source`) are converted to factors to make them easier to plot in subsequent code chunks.
```{r}
join.df <- left_join(ratings.df, old.df, by = c("station", "date", "season", "salzone",  "metric")) %>% 
  filter(date <= max(old.df$date)) %>% 
  unite(index, season, salzone) %>% 
  left_join(unique(bay.df[, c("unique_id", "source")]), by = "unique_id") %>% 
  mutate(index = factor(index),
         metric = factor(metric),
         source = factor(source))
```


```{r}
remove.vec <- c('et4.2_ap_fs1_1985-04-10_spring_m', 'et4.2_ap_fs1_1985-05-07_spring_m', 'et4.2_ap_fs1_1985-05-22_spring_m', 'et4.2_ap_fs1_1985-07-22_summer_m', 'et5.1_ap_fs1_1984-08-23_summer_f', 'et5.2_ap_fs1_1984-08-23_summer_m', 'et5.1_ap_fs1_1984-09-20_summer_o', 'et5.2_ap_fs1_1984-09-20_summer_m', 'et5.2_ap_fs1_1985-04-10_spring_m', 'et5.2_ap_fs1_1985-04-23_spring_m', 'et5.1_ap_fs1_1985-04-24_spring_o', 'et5.1_ap_fs1_1985-05-07_spring_o', 'et5.2_ap_fs1_1985-05-07_spring_m', 'et5.2_ap_fs1_1985-05-21_spring_m', 'et5.1_ap_fs1_1985-05-22_spring_o', 'ee3.1_ap_fs1_1985-04-23_spring_m', 'ee3.1_ap_fs1_1985-05-08_spring_m', 'ee3.1_ap_fs1_1985-05-21_spring_m', 'wt5.1_ap_fs1_1984-09-19_summer_m', 'wt5.1_ap_fs1_1985-04-24_spring_m', 'wt5.1_ap_fs1_1985-05-08_spring_m', 'wt5.1_ap_fs1_1985-05-22_spring_m', 'wt5.1_ap_fs1_1985-07-22_summer_m', 'le2.2_ap_fs1_1984-09-04_summer_m', 'le2.2_ap_fs1_1985-04-15_spring_m', 'ret2.2_ap_fs1_1985-04-15_spring_o', 'ret2.2_ap_fs1_1985-04-30_spring_o', 'le2.2_ap_fs1_1985-04-30_spring_m', 'le2.2_ap_fs1_1985-05-13_spring_m', 'ret2.2_ap_fs1_1985-05-13_spring_o', 'ret2.2_ap_fs1_1985-05-27_spring_o', 'le2.2_ap_fs1_1985-05-27_spring_m', 'ret2.2_ap_fs1_1985-07-08_summer_o', 'le2.2_ap_fs1_1985-07-08_summer_m', 'ret2.2_ap_fs1_1985-07-24_summer_m', 'le2.2_ap_fs1_1985-07-24_summer_m', 'ret2.2_ap_fs1_1985-08-06_summer_m', 'ret2.2_ap_fs1_1985-08-22_summer_o', 'le2.2_ap_fs1_1985-08-22_summer_m', 'cb1.1_ap_fs1_1984-08-07_summer_f', 'cb1.1_ap_fs1_1984-09-26_summer_f', 'cb1.1_ap_fs1_1985-03-21_spring_f', 'cb1.1_ap_fs1_1985-04-10_spring_f', 'cb1.1_ap_fs1_1985-04-24_spring_f', 'cb1.1_ap_fs1_1985-07-10_summer_f', 'tf2.3_ap_fs1_1984-08-06_summer_f', 'tf2.3_ap_fs1_1985-04-15_spring_f', 'tf2.3_ap_fs1_1985-04-30_spring_f', 'tf2.3_ap_fs1_1985-05-13_spring_f', 'tf2.3_ap_fs1_1985-05-27_spring_f', 'tf2.3_ap_fs1_1985-07-08_summer_f', 'tf2.3_ap_fs1_1985-07-24_summer_f', 'tf2.3_ap_fs1_1985-08-06_summer_f', 'tf2.3_ap_fs1_1985-08-22_summer_f', 'tf2.3_ap_fs1_1985-09-09_summer_f', 'le1.1_ap_fs1_1985-03-26_spring_m', 'le1.1_ap_fs1_1985-04-08_spring_m', 'le1.1_ap_fs1_1985-04-22_spring_m', 'tf1.7_ap_fs1_1985-04-22_spring_m', 'le1.1_ap_fs1_1985-05-06_spring_m', 'tf1.7_ap_fs1_1985-05-06_spring_m', 'tf1.5_ap_fs1_1985-05-06_spring_o', 'le1.1_ap_fs1_1985-05-20_spring_m', 'tf1.7_ap_fs1_1985-05-20_spring_m', 'tf1.5_ap_fs1_1985-05-20_spring_f', 'le1.1_ap_fs1_1985-07-17_summer_m', 'cb5.2_ap_fs1_1984-08-06_summer_m', 'cb2.2_ap_fs1_1984-08-07_summer_f', 'cb5.2_ap_fs1_1984-08-29_summer_m', 'cb4.3c_ap_fs1_1984-08-30_summer_m', 'cb3.3c_ap_fs1_1984-08-30_summer_m', 'cb2.2_ap_fs1_1984-08-30_summer_o', 'cb3.3c_ap_fs1_1984-09-11_summer_m', 'cb2.2_ap_fs1_1984-09-12_summer_o', 'cb5.2_ap_fs1_1984-09-24_summer_m', 'cb4.3c_ap_fs1_1984-09-25_summer_m', 'cb3.3c_ap_fs1_1984-09-25_summer_m', 'cb2.2_ap_fs1_1984-09-26_summer_m', 'cb5.2_ap_fs1_1985-03-19_spring_m', 'cb4.3c_ap_fs1_1985-03-20_spring_m', 'cb3.3c_ap_fs1_1985-03-20_spring_m', 'cb5.2_ap_fs1_1985-04-08_spring_m', 'cb4.3c_ap_fs1_1985-04-09_spring_m', 'cb3.3c_ap_fs1_1985-04-09_spring_m', 'cb2.2_ap_fs1_1985-04-10_spring_f', 'cb5.2_ap_fs1_1985-04-22_spring_m', 'cb4.3c_ap_fs1_1985-04-23_spring_m', 'cb3.3c_ap_fs1_1985-04-23_spring_m', 'cb2.2_ap_fs1_1985-04-24_spring_o', 'cb4.3c_ap_fs1_1985-05-07_spring_m', 'cb3.3c_ap_fs1_1985-05-07_spring_m', 'cb3.3c_ap_fs1_1985-05-21_spring_m', 'cb5.2_ap_fs1_1985-07-08_summer_m', 'cb4.3c_ap_fs1_1985-07-09_summer_m', 'cb3.3c_ap_fs1_1985-07-09_summer_m', 'cb2.2_ap_fs1_1985-07-10_summer_o', 'cb5.2_ap_fs1_1985-07-22_summer_p', 'cb3.3c_ap_fs1_1985-07-24_summer_m', 'cb2.2_ap_fs1_1985-07-24_summer_o')

join.df <- join.df %>% 
  filter(!unique_id %in% remove.vec)
```




The numeric values are all rounded to the hundredth place to reduce the detection of minor discrepancies. The absolute difference between the re-calculated IBI score and Jacqueline M. Johnson's IBI score is calculated (`ibi_diff`). Rows are sorted to present IBI score differences in descending order (`ibi_diff`) and only rows where the IBI differences disagree (`ibi_diff > 0`) are retained. This data frame can be used to explore differences in individual metric scores that are causing discrepancies between the two IBI values.
```{r}
join.df <- join.df %>% 
  mutate_if(is.numeric, round, digits = 2) %>% 
  mutate(metric_diff = abs(value - old_metric_value),
         score_diff = abs(score - old_score_value),
         ibi_diff = abs(ibi_score - old_ibi_score)) %>% 
  select(unique_id,  metric, index, source,
         value, old_metric_value, metric_diff,
         score, old_score_value, score_diff,
         ibi_score, old_ibi_score, ibi_diff,
         rating, old_rating)
```

Identify IBI scores calculated in this document that differ from the IBI scores calculated by Jacqueline M. Johnson.
```{r}
diff.df <- join.df %>% 
  filter(ibi_diff > 0) %>% 
  arrange(desc(ibi_diff), desc(score_diff), desc(metric_diff)) %>% 
  distinct()
```

Focus on just the IBI scores that differ. Categorize the IBI score differences (`ibi_diff`) into bins representing 0.5 increments.
```{r}
diff.ibi <- diff.df %>% 
  select(-metric, -value, -old_metric_value, -metric_diff,
         -score, -old_score_value, - score_diff) %>% 
  mutate(disagree_bin = case_when(
    ibi_diff > 0 & ibi_diff <= 0.5 ~ "0 < IBI Score <= 0.5",
    ibi_diff > 0.5 & ibi_diff <= 1 ~ "0.5 < IBI Score <= 1.0",
    ibi_diff > 1 & ibi_diff <= 1.5 ~ "1.0 < IBI Score <= 1.5",
    ibi_diff > 1.5 & ibi_diff <= 2 ~ "1.5 < IBI Score <= 2.0",
    ibi_diff > 2 & ibi_diff <= 2.5 ~ "2.0 < IBI Score <= 2.5",
    ibi_diff > 2.5 & ibi_diff <= 3 ~ "2.5 < IBI Score <= 3.0",
    ibi_diff > 3 & ibi_diff <= 3.5 ~ "3.0 < IBI Score <= 3.5",
    ibi_diff > 3.5 & ibi_diff <= 4 ~ "3.5 < IBI Score <= 4.0",
    TRUE ~ "ERROR"
  )) %>% 
  distinct() %>% 
  arrange(desc(ibi_diff))
```

If at there is at least one IBI score disagreement, then plot the counts of IBI score disagreement for each index (`index`). The bins created in the previous code chunk can be used to divide up the data into separate plots and provide a little bit more information regarding how much the index scores differ.
```{r}
if (nrow(diff.ibi) > 0) {
  ggplot(diff.ibi, aes(index, fill = index)) +
    geom_bar() +
    facet_wrap(~ disagree_bin) + 
    xlab("Index") +
    ylab("Number of Discrepancies") +
    coord_flip() 
}
```

Focus just on scores that differ, which makes it easier to identify errors in this documents code. Subset `disagree.df` to only include metrics (`metric`), metric values (`value`), Jacqueline M. Johnson's scores (`old_score_value`), and this documents scored values (`score`). Find the absolute difference (`score_diff`) between Jacqueline M. Johnson's scores (`old_score_value`) and this documents scored values (`score`). Only rows where the scores disagree (`score_diff > 0`) are retained.
```{r}
diff.score <- diff.df %>% 
  select(unique_id, index, source,  metric, value, old_score_value, score) %>% 
  distinct() %>% 
  mutate(score_diff = abs(old_score_value - score)) %>% 
  filter(score_diff > 0)
```

If there is at least one metric score disagreement, plot the counts of score disagreement for each metric by index (`index`).
```{r, fig.width = 8, fig.height = 6}
if (nrow(diff.score) > 0) {
  ggplot(diff.score, aes(metric, fill = metric)) +
    geom_bar() +
    coord_flip() +
    facet_wrap(~index, ncol = 4)+
    theme(legend.position = "bottom") +
    guides(fill = guide_legend(ncol = 2, bycol = TRUE))
}
```

More information can be gleaned from the previous plot if the data source (`source`) is also used as a grouping factor. Counts of score differences are summed by `metric`, `index`, and `source`.
```{r}
diff.score.source <- diff.score %>% 
  group_by(metric, index, source) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  tidyr::complete(metric, index, source) %>% 
  mutate(count = if_else(is.na(count), as.integer(0), count))
```

If there is at least one metric score disagreement for `diff.score.source`, plot the counts of score disagreement for each metric by source (`source`) and index (`index`).
```{r, fig.width = 8, fig.height = 25}
if (nrow(diff.score.source) > 0) {
  ggplot(diff.score.source, aes(metric, count, fill = metric)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    facet_wrap(~index + source, ncol = 2) +
    theme(legend.position = "bottom") +
    guides(fill = guide_legend(ncol = 2, bycol = TRUE))
}
```




#### Pheophytin Differences

```{r}
pheo.df <- diff.df %>% 
  filter(metric == "pheophytin") %>% 
  separate(unique_id, c("station", "layer", "samplenumber",
                        "sampledate", "season", "salzone"), sep = "_",
           remove = FALSE) %>% 
  mutate(sampledate = as.Date(sampledate))
```

```{r}
pheo.wq <- wq.df %>% 
  filter(str_detect(parameter, "pheo")) %>% 
  rename("sampledate" = "sample_date")
```

```{r eval=FALSE}
search.df <- pmap(list(1:nrow(pheo.df)), function(row.i) {
  pheo.sub <- slice(pheo.df, row.i) %>% 
    mutate_if(is.numeric, round, digits = 2)
  
  wq.sub <- inner_join(pheo.wq, pheo.sub, by = c("station", "sampledate"))
  
  final.df <- wq.sub %>% 
    mutate(one_value = if_else(measurevalue == old_metric_value, TRUE, FALSE),
           mean_wc = if_else(round(mean(wq.sub$measurevalue, na.rm = TRUE), 2) == old_metric_value, TRUE, FALSE),
           mean_surface = if_else(mean(wq.sub[wq.sub$parameter == "s_pheo", "measurevalue"], na.rm = TRUE) == old_metric_value, TRUE, FALSE))
  
  return(final.df)
}) %>% 
  bind_rows() %>% 
  select(unique_id, latinname, size, reportingvalue, dino_abund, old_dino_abund, diff, div)
```


#### Dinoflagellate Biomass Differences

```{r}
dino.df <- bay.taxa %>% 
  select(unique_id) %>% 
  separate(unique_id, c("station", "layer", "samplenumber",
                        "sampledate", "season", "salzone"), sep = "_",
           remove = FALSE) %>% 
  mutate(sampledate = as.Date(sampledate)) %>% 
  distinct() %>% 
  mutate(dino_abund = taxa_abund(bay.taxa, unique_id, reportingvalue, division, "pyrrophycophyta"),
         scrippsiella_precaria_abund = taxa_abund(bay.taxa, unique_id,
                                                    reportingvalue, latinname, "scrippsiella_precaria"),
         dino_abund = dino_abund - scrippsiella_precaria_abund) %>% 
  left_join(unique(bay.df[, c("unique_id", "source")]), by = "unique_id")
```

```{r}
dino.old <- old.org.df %>% 
  mutate(sampledate = as.Date(sample_date)) %>% 
  select(station, sampledate, dino_abund) %>% 
  rename(old_dino_abund = dino_abund) %>% 
  distinct()
```

```{r}
dino.join <- left_join(dino.df, dino.old, by = c("station", "sampledate")) %>% 
  mutate(diff = dino_abund - old_dino_abund,
         div = dino_abund / old_dino_abund) %>% 
  filter(!unique_id %in% remove.vec)
```

```{r}
dino.join %>% 
  filter(diff != 0) %>% 
  unite(index, season, salzone) %>% 
  ggplot(aes(dino_abund, old_dino_abund, color = index)) +
  geom_abline(intercept = 0, slope = 1, color = "green", 
              linetype = "solid", size = 1.5, alpha = 0.1) +
  geom_abline(intercept = 0, slope = 0.5, color = "red", 
              linetype = "solid", size = 1.5, alpha = 0.1) +
  geom_point(alpha = 0.25)
```

```{r}
dino.diff <- dino.join %>% 
  filter(diff != 0,
         div != 2)

search.df <- pmap(list(1:nrow(dino.diff)), function(row.i) {
  dino.sub <- slice(dino.diff, row.i)
  
  raw.sub <- inner_join(taxa.raw, dino.sub, by = c("source", "station", "sampledate", "layer", "samplenumber"))
  
  final.df <- raw.sub %>% 
    filter(reportingvalue == abs(dino.sub$diff))
  
  return(final.df)
}) %>% 
  bind_rows() %>% 
  select(unique_id, latinname, size, reportingvalue, dino_abund, old_dino_abund, diff, div)

search.missing <- search.df %>% 
  filter(diff < 0)

search.remove <- search.df %>% 
  filter(diff > 0)
```

```{r fig.width = 10, fig.height = 10}
search.df %>% 
  separate(unique_id, c("station", "layer", "samplenumber",
                        "date", "season", "salzone"), sep = "_", remove = FALSE) %>% 
  left_join(unique(bay.df[, c("unique_id", "latinname",
                               "source")]), by = c("unique_id", "latinname")) %>% 
  mutate(year = year(date),
         dir = if_else(diff < 0, "too few", "too many")) %>% 
  group_by(latinname, dir) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  mutate(count = if_else(dir == "too few", as.integer(-1 * count), as.integer(count))) %>% 
  group_by(latinname) %>% 
  mutate(total = sum(abs(count))) %>% 
  ungroup() %>% 
  arrange(total) %>% 
  mutate(latinname = factor(latinname, levels = unique(latinname))) %>% 
  ggplot(aes(latinname, count, fill = dir)) +
  scale_fill_manual(values = c("too many" = "#56B4E9",
                               "too few" = "#E69F00")) +
  geom_bar(stat = "identity") +
  coord_flip() #+
#facet_wrap(~year)
```

#### Biomass Differences

```{r eval=FALSE}
mass.df <- bay.taxa %>% 
  select(unique_id) %>% 
  separate(unique_id, c("station", "layer", "samplenumber",
                        "sampledate", "season", "salzone"), sep = "_",
           remove = FALSE) %>% 
  mutate(sampledate = as.Date(sampledate)) %>% 
  distinct() %>% 
  mutate(biomass = taxa_abund(bay.taxa, unique_id, biomass, species)) %>% 
  left_join(unique(bay.df[, c("unique_id", "source")]), by = "unique_id")
```

```{r eval=FALSE}
mass.old <- old.org.df %>% 
  mutate(sampledate = as.Date(sample_date)) %>% 
  select(station, sampledate, tot_biomass) %>% 
  rename(old_biomass = tot_biomass) %>% 
  distinct()
```

```{r eval=FALSE}
mass.join <- left_join(mass.df, mass.old, by = c("station", "sampledate")) %>% 
  mutate(diff = biomass - old_biomass,
         div = biomass / old_biomass)
```

```{r eval=FALSE}
mass.join %>% 
  filter(diff != 0) %>% 
  unite(index, season, salzone) %>% 
  ggplot(aes(biomass, old_biomass, color = index)) +
  geom_abline(intercept = 0, slope = 1, color = "green", 
              linetype = "solid", size = 1.5, alpha = 0.1) +
  geom_abline(intercept = 0, slope = 0.5, color = "red", 
              linetype = "solid", size = 1.5, alpha = 0.1) +
  geom_point(alpha = 0.25)
```


```{r eval=FALSE}
mass.diff <- mass.join %>% 
  filter(abs(diff) > 50)

search.df <- pmap(list(1:nrow(mass.diff)), function(row.i) {
  mass.sub <- slice(mass.diff, row.i)
  
  raw.sub <- inner_join(taxa.raw, mass.sub, by = c("source", "station", "sampledate", "layer", "samplenumber"))
  
  final.df <- raw.sub %>% 
    filter(biomass == abs(mass.sub$diff))
  
  return(final.df)
}) %>% 
  bind_rows() %>% 
  select(unique_id, latinname, size, reportingvalue, biomass, old_biomass, diff, div)

search.missing <- search.df %>% 
  filter(diff < 0)

search.remove <- search.df %>% 
  filter(diff > 0)
```

```{r fig.width = 10, fig.height = 25, eval=FALSE}
search.df %>% 
  separate(unique_id, c("station", "layer", "samplenumber", "date", "season", "salzone"), sep = "_", remove = FALSE) %>% 
  left_join(unique(bay.df[, c("unique_id", "latinname", "size", "source")]), by = c("unique_id", "latinname")) %>% 
  mutate(year = year(date),
         dir = if_else(diff < 0, "neg", "pos")) %>% 
  group_by(latinname, dir) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  mutate(count = if_else(dir == "neg", as.integer(-1 * count), as.integer(count))) %>% 
  group_by(latinname) %>% 
  mutate(total = sum(abs(count))) %>% 
  ungroup() %>% 
  arrange(total) %>% 
  mutate(latinname = factor(latinname, levels = unique(latinname))) %>% 
  ggplot(aes(latinname, count, fill = dir)) +
  scale_fill_manual(values = c("pos" = "#56B4E9",
                               "neg" = "#E69F00")) +
  geom_bar(stat = "identity") +
  coord_flip() #+
#facet_wrap(~year)
```

#### Abundance Differences

The ratio of total phytoplankton biomass to chlorophyll a is frequently different from the values calculated by Jacqueline M. Johnson. The biomass values are calculated from the reported taxonomic counts (`reportingvalue`); therefore, I decided to explore if there were differences between the abundance values calculated in this document and the abundance values calculated by Jacqueline M. Johnson.


Using `bay.taxa` from the [Metric Calculation] section, the taxonomic abundances are found for each sampling event (`unique_id`).
```{r}
abund.df <- bay.taxa %>% 
  filter(!unique_id %in% remove.vec) %>% 
  select(unique_id) %>% 
  separate(unique_id, c("station", "layer", "samplenumber",
                        "sampledate", "season", "salzone"), sep = "_",
           remove = FALSE) %>% 
  mutate(sampledate = as.Date(sampledate)) %>% 
  distinct() %>% 
  mutate(abundance = taxa_abund(bay.taxa[!is.na(bay.taxa$biomass) & !(bay.taxa$unique_id %in% remove.vec), ], unique_id, reportingvalue, species)) %>% 
  left_join(unique(bay.df[, c("unique_id", "source")]), by = "unique_id")
```

Abundance values calculated by Jacqueline M. Johnson are extracted from `old.org.df` (created in the [Import Old Values] section).
```{r}
abund.old <- old.org.df %>% 
  mutate(sampledate = as.Date(sample_date)) %>% 
  select(station, sampledate, tot_abund) %>% 
  rename(old_abund = tot_abund) %>% 
  distinct()
```

The two independent abundance calculations are joined into a single data frame and the difference between the two values is calculated (`diff`). Also, a subset of abundances calculated in this document were found to be exactly double the abundances calculated by Jacqueline M. Johnson; to further explore this phenomena, the abundance calculated in this document was divided by the abunance calculated by Jacqueline M. Johnson (`div`).
```{r}
abund.join <- left_join(abund.df, abund.old, by = c("station", "sampledate")) %>% 
  mutate(diff = abundance - old_abund,
         div = abundance / old_abund)
```

`r paste0(round(((abund.join %>% filter(diff != 0) %>% nrow()) / (abund.join %>% nrow())) * 100, 1), "%")` of the sampling events had no difference between the abundance values. The values in agreement are excluded to observe patterns associated with values that differ from one another. Also there was one extreme point represented by an abundance greater than 2e+09, which was excluded becuase it was obsuring the view of the remaining points. 
```{r}
abund.join %>% 
  filter(abundance < 2e+09,
         diff != 0) %>% 
  unite(index, season, salzone) %>% 
  ggplot(aes(abundance, old_abund, color = index)) +
  geom_abline(intercept = 0, slope = 1, color = "green", 
              linetype = "solid", size = 1.5, alpha = 0.1) +
  geom_abline(intercept = 0, slope = 0.5, color = "red", 
              linetype = "solid", size = 1.5, alpha = 0.1) +
  geom_point(alpha = 0.25)
```

The same plot is divided into seperate plots representing each index and data source.
```{r, fig.width = 10, fig.height = 15}
abund.join %>% 
  filter(abundance < 2e+09,
         diff != 0) %>% 
  unite(index, season, salzone) %>% 
  ggplot(aes(abundance, old_abund, color = index)) +
  geom_point(alpha = 0.25) +
  facet_wrap(~index + source, ncol = 2) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(ncol = 2, bycol = TRUE)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


```{r}
test <- abund.join %>% 
  filter(source == "msu/pearl",
         season == "summer",
         salzone == "m") %>% 
  mutate(year = year(sampledate))
```

```{r, fig.width = 10, fig.height = 25}
test %>% 
  ggplot(aes(abundance, old_abund, color = samplenumber)) +
  geom_point(alpha = 0.25) +
  facet_wrap(~year, ncol = 2, scale = "free") +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(ncol = 2, bycol = TRUE)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```





```{r}
abund.join %>% 
  filter(diff != 0,
         div != 2) %>% 
  unite(index, season, salzone) %>% 
  ggplot(aes(sampledate, diff, color = source)) +
  geom_point(alpha = 0.25)
```


```{r}
abund.diff <- abund.join %>% 
  filter(diff != 0)

search.df <- pmap(list(1:nrow(abund.diff)), function(row.i) {
  abund.sub <- slice(abund.diff, row.i)
  
  raw.sub <- inner_join(taxa.raw, abund.sub, by = c("source", "station", "sampledate", "layer", "samplenumber"))
  
  final.df <- raw.sub %>% 
    filter(reportingvalue == abs(abund.sub$diff))
  
  return(final.df)
}) %>% 
  bind_rows() %>% 
  select(unique_id, latinname, size, reportingvalue, abundance, old_abund, diff, div)

search.missing <- search.df %>% 
  filter(diff < 0)

search.remove <- search.df %>% 
  filter(diff > 0)
```

```{r fig.width = 10, fig.height = 25}
search.df %>% 
  separate(unique_id, c("station", "layer", "samplenumber", "date", "season", "salzone"), sep = "_", remove = FALSE) %>% 
  left_join(unique(bay.df[, c("unique_id", "latinname", "size", "source")]), by = c("unique_id", "latinname")) %>% 
  mutate(year = year(date),
         dir = if_else(diff < 0, "too few", "too many")) %>% 
  group_by(latinname, dir) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  mutate(count = if_else(dir == "too few", as.integer(-1 * count), as.integer(count))) %>% 
  group_by(latinname) %>% 
  mutate(total = sum(abs(count))) %>% 
  ungroup() %>% 
  arrange(total) %>% 
  mutate(latinname = factor(latinname, levels = unique(latinname))) %>% 
  ggplot(aes(latinname, count, fill = dir)) +
  scale_fill_manual(values = c("too many" = "#56B4E9",
                               "too few" = "#E69F00")) +
  geom_bar(stat = "identity") +
  coord_flip() #+
#facet_wrap(~year)
```



```{r fig.width = 10, fig.height = 7}
search.missing %>% 
  separate(unique_id, c("station", "layer", "samplenumber", "date", "season", "salzone"), sep = "_", remove = FALSE) %>% 
  left_join(unique(bay.df[, c("unique_id", "source")]), by = "unique_id") %>% 
  mutate(year = year(date)) %>% 
  group_by(year, source, latinname) %>% 
  summarize(count = n()) %>% 
  ggplot(aes(year, count, fill = source)) +
  scale_fill_manual(values = c("odu/pel" = "#56B4E9",
                               "msu/pearl" = "#E69F00")) +
  geom_bar(stat = "identity") +
  facet_wrap(~latinname + source, ncol = 5)
```

```{r fig.width = 10, fig.height = 20}
search.remove %>% 
  separate(unique_id, c("station", "layer", "samplenumber", "date", "season", "salzone"), sep = "_", remove = FALSE) %>% 
  left_join(unique(bay.df[, c("unique_id", "source")]), by = "unique_id") %>% 
  mutate(year = year(date)) %>% 
  group_by(year, source, latinname) %>% 
  summarize(count = n()) %>% 
  ggplot(aes(year, count, fill = source)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("odu/pel" = "#56B4E9",
                               "msu/pearl" = "#E69F00")) +
  facet_wrap(~latinname + source, ncol = 5)
```

## Conclusion

Currently, there are a large number of descrepancies between the IBI values calculated in this document and the IBI values found by Jacqueline M. Johnson of the Chesapeake Bay Program. The scoring functions provided in this document appear to be functioning correctly based on the [Scoring Disagreement] section. After using the scoring functions to re-score the metric values provided by Jacqueline M. Johnson, several errors were identified and corrected. These corrections reducing the scoring descrepancies to less than 10. Most of the remaining descrepancies are found at scoring thresholds. Jacqueline M. Johnson's code provided the best description of scoring thresholds, which have been replicated in the [Scoring Functions] section; at this time, I have no justification for altering the thresholds further, so the few descrapencies will remain. Ultimately, the [Scoring Disagreement] section suggests that any remaining descrapancies between the values found in this document and the values values found by Jacqueline M. Johnson are caused by differences in metric calculations and/or, more likely, differences in taxonomic count preperations. So far, no table has been found representing Jacqueline M. Johnson's prepared taxonomic counts. This table would make it easier to identify descrepancies in how the taxa are prepared (e.g. excluded taxa or assigned carbon values). Without this table I need help identifying why these descrepancies exist. If you have any suggestions, please let me know.