---
title: ''
output: html_document
---

#### Pheophytin Differences

```{r}
pheo.df <- diff.df %>% 
  filter(metric == "pheophytin") %>% 
  separate(unique_id, c("station", "layer", "samplenumber",
                        "sampledate", "season", "salzone"), sep = "_",
           remove = FALSE) %>% 
  mutate(sampledate = as.Date(sampledate))
```

```{r}
pheo.wq <- wq.df %>% 
  filter(str_detect(parameter, "pheo")) %>% 
  rename("sampledate" = "sample_date")
```

```{r eval=FALSE}
search.df <- pmap(list(1:nrow(pheo.df)), function(row.i) {
  pheo.sub <- slice(pheo.df, row.i) %>% 
    mutate_if(is.numeric, round, digits = 2)
  
  wq.sub <- inner_join(pheo.wq, pheo.sub, by = c("station", "sampledate"))
  
  final.df <- wq.sub %>% 
    mutate(one_value = if_else(measurevalue == old_metric_value, TRUE, FALSE),
           mean_wc = if_else(round(mean(wq.sub$measurevalue, na.rm = TRUE), 2) == old_metric_value, TRUE, FALSE),
           mean_surface = if_else(mean(wq.sub[wq.sub$parameter == "s_pheo", "measurevalue"], na.rm = TRUE) == old_metric_value, TRUE, FALSE))
  
  return(final.df)
}) %>% 
  bind_rows() %>% 
  select(unique_id, latinname, size, reportingvalue, dino_abund, old_dino_abund, diff, div)
```


#### Dinoflagellate Biomass Differences

```{r}
dino.df <- bay.taxa %>% 
  select(unique_id) %>% 
  separate(unique_id, c("station", "layer", "samplenumber",
                        "sampledate", "season", "salzone"), sep = "_",
           remove = FALSE) %>% 
  mutate(sampledate = as.Date(sampledate)) %>% 
  distinct() %>% 
  mutate(dino_abund = taxa_abund(bay.taxa, unique_id, reportingvalue, division, "pyrrophycophyta"),
         scrippsiella_precaria_abund = taxa_abund(bay.taxa, unique_id,
                                                  reportingvalue, latinname, "scrippsiella_precaria"),
         dino_abund = dino_abund - scrippsiella_precaria_abund) %>% 
  left_join(unique(bay.df[, c("unique_id", "source")]), by = "unique_id")
```

```{r}
dino.old <- old.org.df %>% 
  mutate(sampledate = as.Date(sample_date)) %>% 
  select(station, sampledate, dino_abund) %>% 
  rename(old_dino_abund = dino_abund) %>% 
  distinct()
```

```{r}
dino.join <- left_join(dino.df, dino.old, by = c("station", "sampledate")) %>% 
  mutate(diff = dino_abund - old_dino_abund,
         div = dino_abund / old_dino_abund) %>% 
  filter(!unique_id %in% remove.vec)
```

```{r}
dino.join %>% 
  filter(diff != 0) %>% 
  unite(index, season, salzone) %>% 
  ggplot(aes(dino_abund, old_dino_abund, color = index)) +
  geom_abline(intercept = 0, slope = 1, color = "green", 
              linetype = "solid", size = 1.5, alpha = 0.1) +
  geom_abline(intercept = 0, slope = 0.5, color = "red", 
              linetype = "solid", size = 1.5, alpha = 0.1) +
  geom_point(alpha = 0.25)
```

```{r}
dino.diff <- dino.join %>% 
  filter(diff != 0,
         div != 2)

search.df <- pmap(list(1:nrow(dino.diff)), function(row.i) {
  dino.sub <- slice(dino.diff, row.i)
  
  raw.sub <- inner_join(taxa.raw, dino.sub, by = c("source", "station", "sampledate", "layer", "samplenumber"))
  
  final.df <- raw.sub %>% 
    filter(reportingvalue == abs(dino.sub$diff))
  
  return(final.df)
}) %>% 
  bind_rows() %>% 
  select(unique_id, latinname, size, reportingvalue, dino_abund, old_dino_abund, diff, div)

search.missing <- search.df %>% 
  filter(diff < 0)

search.remove <- search.df %>% 
  filter(diff > 0)
```

```{r fig.width = 10, fig.height = 10}
search.df %>% 
  separate(unique_id, c("station", "layer", "samplenumber",
                        "date", "season", "salzone"), sep = "_", remove = FALSE) %>% 
  left_join(unique(bay.df[, c("unique_id", "latinname",
                              "source")]), by = c("unique_id", "latinname")) %>% 
  mutate(year = year(date),
         dir = if_else(diff < 0, "too few", "too many")) %>% 
  group_by(latinname, dir) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  mutate(count = if_else(dir == "too few", as.integer(-1 * count), as.integer(count))) %>% 
  group_by(latinname) %>% 
  mutate(total = sum(abs(count))) %>% 
  ungroup() %>% 
  arrange(total) %>% 
  mutate(latinname = factor(latinname, levels = unique(latinname))) %>% 
  ggplot(aes(latinname, count, fill = dir)) +
  scale_fill_manual(values = c("too many" = "#56B4E9",
                               "too few" = "#E69F00")) +
  geom_bar(stat = "identity") +
  coord_flip() #+
#facet_wrap(~year)
```

#### Biomass Differences

```{r eval=FALSE}
mass.df <- bay.taxa %>% 
  select(unique_id) %>% 
  separate(unique_id, c("station", "layer", "samplenumber",
                        "sampledate", "season", "salzone"), sep = "_",
           remove = FALSE) %>% 
  mutate(sampledate = as.Date(sampledate)) %>% 
  distinct() %>% 
  mutate(biomass = taxa_abund(bay.taxa, unique_id, biomass, species)) %>% 
  left_join(unique(bay.df[, c("unique_id", "source")]), by = "unique_id")
```

```{r eval=FALSE}
mass.old <- old.org.df %>% 
  mutate(sampledate = as.Date(sample_date)) %>% 
  select(station, sampledate, tot_biomass) %>% 
  rename(old_biomass = tot_biomass) %>% 
  distinct()
```

```{r eval=FALSE}
mass.join <- left_join(mass.df, mass.old, by = c("station", "sampledate")) %>% 
  mutate(diff = biomass - old_biomass,
         div = biomass / old_biomass)
```

```{r eval=FALSE}
mass.join %>% 
  filter(diff != 0) %>% 
  unite(index, season, salzone) %>% 
  ggplot(aes(biomass, old_biomass, color = index)) +
  geom_abline(intercept = 0, slope = 1, color = "green", 
              linetype = "solid", size = 1.5, alpha = 0.1) +
  geom_abline(intercept = 0, slope = 0.5, color = "red", 
              linetype = "solid", size = 1.5, alpha = 0.1) +
  geom_point(alpha = 0.25)
```


```{r eval=FALSE}
mass.diff <- mass.join %>% 
  filter(abs(diff) > 50)

search.df <- pmap(list(1:nrow(mass.diff)), function(row.i) {
  mass.sub <- slice(mass.diff, row.i)
  
  raw.sub <- inner_join(taxa.raw, mass.sub, by = c("source", "station", "sampledate", "layer", "samplenumber"))
  
  final.df <- raw.sub %>% 
    filter(biomass == abs(mass.sub$diff))
  
  return(final.df)
}) %>% 
  bind_rows() %>% 
  select(unique_id, latinname, size, reportingvalue, biomass, old_biomass, diff, div)

search.missing <- search.df %>% 
  filter(diff < 0)

search.remove <- search.df %>% 
  filter(diff > 0)
```

```{r fig.width = 10, fig.height = 25, eval=FALSE}
search.df %>% 
  separate(unique_id, c("station", "layer", "samplenumber", "date", "season", "salzone"), sep = "_", remove = FALSE) %>% 
  left_join(unique(bay.df[, c("unique_id", "latinname", "size", "source")]), by = c("unique_id", "latinname")) %>% 
  mutate(year = year(date),
         dir = if_else(diff < 0, "neg", "pos")) %>% 
  group_by(latinname, dir) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  mutate(count = if_else(dir == "neg", as.integer(-1 * count), as.integer(count))) %>% 
  group_by(latinname) %>% 
  mutate(total = sum(abs(count))) %>% 
  ungroup() %>% 
  arrange(total) %>% 
  mutate(latinname = factor(latinname, levels = unique(latinname))) %>% 
  ggplot(aes(latinname, count, fill = dir)) +
  scale_fill_manual(values = c("pos" = "#56B4E9",
                               "neg" = "#E69F00")) +
  geom_bar(stat = "identity") +
  coord_flip() #+
#facet_wrap(~year)
```

#### Abundance Differences

The ratio of total phytoplankton biomass to chlorophyll a is frequently different from the values calculated by Jacqueline M. Johnson. The biomass values are calculated from the reported taxonomic counts (`reportingvalue`); therefore, I decided to explore if there were differences between the abundance values calculated in this document and the abundance values calculated by Jacqueline M. Johnson.


Using `bay.taxa` from the [Metric Calculation] section, the taxonomic abundances are found for each sampling event (`unique_id`).
```{r}
abund.df <- bay.taxa %>% 
  filter(!unique_id %in% remove.vec) %>% 
  select(unique_id) %>% 
  separate(unique_id, c("station", "layer", "samplenumber",
                        "sampledate", "season", "salzone"), sep = "_",
           remove = FALSE) %>% 
  mutate(sampledate = as.Date(sampledate)) %>% 
  distinct() %>% 
  mutate(abundance = taxa_abund(bay.taxa[!is.na(bay.taxa$biomass) & !(bay.taxa$unique_id %in% remove.vec), ], unique_id, reportingvalue, species)) %>% 
  left_join(unique(bay.df[, c("unique_id", "source")]), by = "unique_id")
```

Abundance values calculated by Jacqueline M. Johnson are extracted from `old.org.df` (created in the [Import Old Values] section).
```{r}
abund.old <- old.org.df %>% 
  mutate(sampledate = as.Date(sample_date)) %>% 
  select(station, sampledate, tot_abund) %>% 
  rename(old_abund = tot_abund) %>% 
  distinct()
```

The two independent abundance calculations are joined into a single data frame and the difference between the two values is calculated (`diff`). Also, a subset of abundances calculated in this document were found to be exactly double the abundances calculated by Jacqueline M. Johnson; to further explore this phenomena, the abundance calculated in this document was divided by the abundance calculated by Jacqueline M. Johnson (`div`).
```{r}
abund.join <- left_join(abund.df, abund.old, by = c("station", "sampledate")) %>% 
  mutate(diff = abundance - old_abund,
         div = abundance / old_abund)
```

`r paste0(round(((abund.join %>% filter(diff != 0) %>% nrow()) / (abund.join %>% nrow())) * 100, 1), "%")` of the sampling events had no difference between the abundance values. The values in agreement are excluded to observe patterns associated with values that differ from one another. Also there was one extreme point represented by an abundance greater than 2e+09, which was excluded because it was obscuring the view of the remaining points. 
```{r}
abund.join %>% 
  filter(abundance < 2e+09,
         diff != 0) %>% 
  unite(index, season, salzone) %>% 
  ggplot(aes(abundance, old_abund, color = index)) +
  geom_abline(intercept = 0, slope = 1, color = "green", 
              linetype = "solid", size = 1.5, alpha = 0.1) +
  geom_abline(intercept = 0, slope = 0.5, color = "red", 
              linetype = "solid", size = 1.5, alpha = 0.1) +
  geom_point(alpha = 0.25)
```

The same plot is divided into separate plots representing each index and data source.
```{r, fig.width = 10, fig.height = 15}
abund.join %>% 
  filter(abundance < 2e+09,
         diff != 0) %>% 
  unite(index, season, salzone) %>% 
  ggplot(aes(abundance, old_abund, color = index)) +
  geom_point(alpha = 0.25) +
  facet_wrap(~index + source, ncol = 2) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(ncol = 2, bycol = TRUE)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


```{r}
test <- abund.join %>% 
  filter(source == "msu/pearl",
         season == "summer",
         salzone == "m") %>% 
  mutate(year = year(sampledate))
```

```{r, fig.width = 10, fig.height = 25}
test %>% 
  ggplot(aes(abundance, old_abund, color = samplenumber)) +
  geom_point(alpha = 0.25) +
  facet_wrap(~year, ncol = 2, scale = "free") +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(ncol = 2, bycol = TRUE)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```





```{r}
abund.join %>% 
  filter(diff != 0,
         div != 2) %>% 
  unite(index, season, salzone) %>% 
  ggplot(aes(sampledate, diff, color = source)) +
  geom_point(alpha = 0.25)
```


```{r}
abund.diff <- abund.join %>% 
  filter(diff != 0)

search.df <- pmap(list(1:nrow(abund.diff)), function(row.i) {
  abund.sub <- slice(abund.diff, row.i)
  
  raw.sub <- inner_join(taxa.raw, abund.sub, by = c("source", "station", "sampledate", "layer", "samplenumber"))
  
  final.df <- raw.sub %>% 
    filter(reportingvalue == abs(abund.sub$diff))
  
  return(final.df)
}) %>% 
  bind_rows() %>% 
  select(unique_id, latinname, size, reportingvalue, abundance, old_abund, diff, div)

search.missing <- search.df %>% 
  filter(diff < 0)

search.remove <- search.df %>% 
  filter(diff > 0)
```

```{r fig.width = 10, fig.height = 25}
search.df %>% 
  separate(unique_id, c("station", "layer", "samplenumber", "date", "season", "salzone"), sep = "_", remove = FALSE) %>% 
  left_join(unique(bay.df[, c("unique_id", "latinname", "size", "source")]), by = c("unique_id", "latinname")) %>% 
  mutate(year = year(date),
         dir = if_else(diff < 0, "too few", "too many")) %>% 
  group_by(latinname, dir) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  mutate(count = if_else(dir == "too few", as.integer(-1 * count), as.integer(count))) %>% 
  group_by(latinname) %>% 
  mutate(total = sum(abs(count))) %>% 
  ungroup() %>% 
  arrange(total) %>% 
  mutate(latinname = factor(latinname, levels = unique(latinname))) %>% 
  ggplot(aes(latinname, count, fill = dir)) +
  scale_fill_manual(values = c("too many" = "#56B4E9",
                               "too few" = "#E69F00")) +
  geom_bar(stat = "identity") +
  coord_flip() #+
#facet_wrap(~year)
```



```{r fig.width = 10, fig.height = 7}
search.missing %>% 
  separate(unique_id, c("station", "layer", "samplenumber", "date", "season", "salzone"), sep = "_", remove = FALSE) %>% 
  left_join(unique(bay.df[, c("unique_id", "source")]), by = "unique_id") %>% 
  mutate(year = year(date)) %>% 
  group_by(year, source, latinname) %>% 
  summarize(count = n()) %>% 
  ggplot(aes(year, count, fill = source)) +
  scale_fill_manual(values = c("odu/pel" = "#56B4E9",
                               "msu/pearl" = "#E69F00")) +
  geom_bar(stat = "identity") +
  facet_wrap(~latinname + source, ncol = 5)
```

```{r fig.width = 10, fig.height = 20}
search.remove %>% 
  separate(unique_id, c("station", "layer", "samplenumber", "date", "season", "salzone"), sep = "_", remove = FALSE) %>% 
  left_join(unique(bay.df[, c("unique_id", "source")]), by = "unique_id") %>% 
  mutate(year = year(date)) %>% 
  group_by(year, source, latinname) %>% 
  summarize(count = n()) %>% 
  ggplot(aes(year, count, fill = source)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("odu/pel" = "#56B4E9",
                               "msu/pearl" = "#E69F00")) +
  facet_wrap(~latinname + source, ncol = 5)
```
