---
title: "Map Ratings"
output: html_document
---
```{r echo=FALSE}
knitr::opts_chunk$set(eval=evaluate, cache=cache.me)
```

Summarize `ratings.df` to just represent IBI scores and ratings. Join `ratings.df` with `stations.df` to provide latitude and longitude data for interactive maps. `jitter()` is used to alter the points position slightly, to prevent event data for the same station from completely overlapping one another.
```{r}
ibi.df <- ratings.df %>% 
  select(-metric, -value, -score) %>% 
  unite(season_sal, season, salzone) %>% 
  distinct() %>% 
  left_join(stations.df, by = "station") %>% 
  mutate(longitude = jitter(longitude, amount = 0.01),
         latitude = jitter(latitude, amount = 0.01))
```

Convert the `rating` column to factors, which will make it easier to assign colors to the ratings in subsequent plots.
```{r}
ibi.df <- ibi.df %>% 
  mutate(rating = factor(rating, c("poor", "fair_poor", "fair", "fair_good", "good")))
```

A color palette is assigned to the ratings.
```{r}
factpal <- colorFactor(c("red", "orange", "yellow", "lightgreen", "green"), ibi.df$rating)
```

`leaflet` is used to provide a visual reference and check of the ratings.
```{r}
leaflet(ibi.df) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>%  
  addCircleMarkers( ~longitude, ~latitude,
                    color = ~factpal(rating),
                    stroke = FALSE,
                    fillOpacity = 0.5,
                    popup = paste("Station:", stations.df$station, "<br/>",
                                  "Agency:", stations.df$agency, "<br/>",
                                  "Source:", stations.df$datasource, "<br/>",
                                  "Latitude:", stations.df$latitude, "<br/>",
                                  "Longitude:", stations.df$longitude))
```

Plot each stations IBI trend for each applicable index. The background colors represent the five rating category ranges (i.e. "poor", "fair_poor", "fair", "fair_good", "good"), providing the user with a quick reference of which rating a particular point would receive.
```{r, fig.width = 12, fig.height = 12, message=FALSE}
min.date <- min(ibi.df$date, na.rm = TRUE)
max.date <- max(ibi.df$date, na.rm = TRUE)

ggplot(data = ibi.df, aes(date, ibi_score, group = season_sal, color = season_sal)) +
  theme_classic() +
  scale_x_date(limits = c(min(ibi.df$date, na.rm = TRUE), max(ibi.df$date, na.rm = TRUE)), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0.8, 5.2), expand = c(0, 0)) +
  annotate("rect", xmin = min.date, xmax = max.date, ymin = 0.8, ymax = 2, fill = "red", alpha = 0.25) +
  annotate("rect", xmin = min.date, xmax = max.date, ymin = 2, ymax = 2.67, fill = "orange", alpha = 0.25) +
  annotate("rect", xmin = min.date, xmax = max.date, ymin = 2.67, ymax = 3.33, fill = "yellow", alpha = 0.25) +
  annotate("rect", xmin = min.date, xmax = max.date, ymin = 3.33, ymax = 4, fill = "lightgreen", alpha = 0.25) +
  annotate("rect", xmin = min.date, xmax = max.date, ymin = 4, ymax = 5.2, fill = "green", alpha = 0.25) +
  geom_line() +
  geom_point() +
  facet_wrap(~ station, ncol = 5, scales = "free")

```

These figures are the same as the previous figures but the plots have been divided by index (e.g. Spring Freshwater and Summer Polyhaline) to reduce clutter.
```{r, fig.width = 12, fig.height = 60, message=FALSE}
ggplot(data = ibi.df, aes(date, ibi_score, group = season_sal, color = season_sal)) +
  theme_classic() +
  scale_x_date(limits = c(min(ibi.df$date, na.rm = TRUE), max(ibi.df$date, na.rm = TRUE)), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0.8, 5.2), expand = c(0, 0)) +
  annotate("rect", xmin = min.date, xmax = max.date, ymin = 0.8, ymax = 2, fill = "red", alpha = 0.25) +
  annotate("rect", xmin = min.date, xmax = max.date, ymin = 2, ymax = 2.67, fill = "orange", alpha = 0.25) +
  annotate("rect", xmin = min.date, xmax = max.date, ymin = 2.67, ymax = 3.33, fill = "yellow", alpha = 0.25) +
  annotate("rect", xmin = min.date, xmax = max.date, ymin = 3.33, ymax = 4, fill = "lightgreen", alpha = 0.25) +
  annotate("rect", xmin = min.date, xmax = max.date, ymin = 4, ymax = 5.2, fill = "green", alpha = 0.25) +
  geom_line(na.rm = TRUE) +
  geom_point(na.rm = TRUE) +
  facet_wrap(~ station + season_sal, ncol = 5, scales = "free")

```

Another way of looking at the data above is represent all of the samples for a particular station in a box-and-whisker plot. This provides an a good summary of the scores for each station. However, this plot could be miss leading because the majority of stations have frequencies and provides no frame of reference for changes overtime.
```{r, fig.width = 12, fig.height = 12}

station.vec <- sort(unique(ibi.df$station))
min.station <- station.vec[1]
max.station <- station.vec[length(station.vec)]
ggplot(data = ibi.df, aes(station, ibi_score)) +
  theme_classic() +
  #scale_x_discrete(expand = c(0, 0)) +
  scale_y_continuous(limits = c(0.8, 5.2), expand = c(0, 0)) +
  annotate("rect", xmin = min.station, xmax = max.station, ymin = 0.8, ymax = 2, fill = "red", alpha = 0.25) +
  annotate("rect", xmin = min.station, xmax = max.station, ymin = 2, ymax = 2.67, fill = "orange", alpha = 0.25) +
  annotate("rect", xmin = min.station, xmax = max.station, ymin = 2.67, ymax = 3.33, fill = "yellow", alpha = 0.25) +
  annotate("rect", xmin = min.station, xmax = max.station, ymin = 3.33, ymax = 4, fill = "lightgreen", alpha = 0.25) +
  annotate("rect", xmin = min.station, xmax = max.station, ymin = 4, ymax = 5.2, fill = "green", alpha = 0.25) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  facet_wrap(~season_sal, ncol = 2)
```

