---
title: "Hierarchy"
output: html_document
---
```{r echo=FALSE}
knitr::opts_chunk$set(eval=evaluate, cache=cache.me)
```

This section joins the phytoplankton data obtained from CEDR ([CEDR Data]) and the taxonomic hierarcy information data obtained from ITIS ([ITIS Data]). 

### Taxonomic Data

Phytoplankton taxonomic count data obtained from CEDR ([Download Phytoplankton Data]) is imported.
```{r}
col.class.vec <- c("samplenumber" = "character",
                   "tsn" = "character",
                   "speccode" = "character",
                   "reportingvalue" = "integer")

taxa.raw <- data.table::fread(file.path(project.dir, "data/phytoplankton/cedr_phyto_taxa.csv"),
                            data.table = FALSE,
                            colClasses = col.class.vec,
                            na.strings = "") %>% 
  mutate(sampledate = as.Date(sampledate))
```

The PIBI was developed for the taxa found above the pycnocline. Therefore, rows where the layer column does not specifies above the pycnocline ("ap") or water column ("wc") are removed. Furthermore, [@LacouturePhytoplanktonindexbiotic2006] (page 599) developed indices for data collected in the spring (March-May) and summer (July-September).
```{r}
bay.df <- taxa.raw %>% 
  filter(layer %in% c("ap", "wc")) %>% 
  distinct() %>% 
  mutate(month = month(sampledate),
         season = case_when(
           month %in% c(3, 4, 5) ~ "spring",
           month %in% c(7, 8, 9) ~ "summer",
           TRUE ~ "remove"
         )) %>% 
  filter(season %in% c("spring", "summer"))
```

If there are multiple taxa with the same taxonomic name, then the reporting values should be summed. Each row should represent a unique taxon.
```{r}
bay.df <- bay.df %>% 
  group_by_at(vars(-reportingvalue)) %>% 
  summarize(reportingvalue = sum(reportingvalue)) %>% 
  ungroup()
```

Only phytoplankton data ("ph") is retained for analysis. Additionally, some taxa were not identified to any useful taxonomic rank and were excluded from the analysis. 
```{r}
bay.df <- bay.df %>% 
  # Removes methods with prefix "BE" (Tidal Benthic Taxa Enumeration) or blanks.
  filter(grepl("ph", method),
         !latinname %in% c("micro-phytoflagellates",
                           "microflagellates",
                           #"green_cells",
                           #"blue_green_sphere",
                           "epiphytic_flagellates",
                           "hydrodictyon_reticulatum"))
```


#### Taxonomic Hierarchy

Taxonomic hierarchy data obtained from ITIS ([ITIS Data]) is imported.
```{r}
hier.wide <- data.table::fread(file.path(project.dir, "data/itis/itis_hierarchy.csv"),
                            data.table = FALSE,
                            na.strings = "") %>% 
  clean_up()
```

The taxonomic counts are joined with the taxonomic hierarchy data by the `final_tsn` columns. A unique identifier (`unique_id`) is created for each sampling event by concatenating `station`, `layer`, `samplenumber`, and `sampledate`.
```{r}
bay.df <- left_join(bay.df, hier.wide, by = c("final_tsn" = "org_tsn")) %>% 
  mutate(unique_id = paste(station, layer, samplenumber, sampledate, sep = "_"))
```

Some taxa require specific attention before they can be finalized.
```{r}
bay.df <- bay.df %>% 
  mutate(
    final_id = if_else(latinname == "trinacria_regina", "trinacria_regina",  final_id),
    final_id = case_when(
      latinname == "green_cells" ~ "green_cells",
      latinname == "blue_green_spheres" ~ "blue_green_spheres",
      TRUE ~ final_id
    ),
    kingdom = if_else(latinname == "trinacria_regina",
                      "chromista", kingdom),
    phylum = if_else(latinname == "trinacria_regina",
                     "bacillariophyta", phylum),
    subphylum = if_else(latinname == "trinacria_regina",
                        "bacillariophytina", subphylum),
    class = if_else(latinname == "trinacria_regina",
                    "mediophyceae", class),
    subclass = if_else(latinname == "trinacria_regina",
                       "chaetocerotophycidae", subclass),
    order = if_else(latinname == "trinacria_regina",
                    "hemiaulales", order),
    family = if_else(latinname == "trinacria_regina",
                     "hemiaulaceae", family),
    genus = case_when(latinname == "trinacria_regina" ~ "trinacria",
                      latinname == "didymocystis" ~ "didymocystis",
                      latinname == "Lauterborniella_elegantissima" ~ "lauterborniella",
                      TRUE ~ genus),
    species = case_when(latinname == "trinacria_regina" ~ "trinacria_regina",
                        latinname == "lauterborniella_elegantissima" ~ "lauterborniella_elegantissima",
                        TRUE ~ species)
  )
```

Remove all unnecessary objects from the environment.
```{r}
rm(col.class.vec)
```

```{r}
test <- bay.df %>% 
  filter(is.na(kingdom))
```

