---
title: "R Notebook"
output: html_document
---
```{r echo=FALSE}
knitr::opts_chunk$set(eval=evaluate, cache=cache.me)
```

Phytoplankton sampling event data obtained from CEDR ([Download Monitoring Event]) is imported.
```{r}
events.df <- data.table::fread(file.path(project.dir, "data/phytoplankton/cedr_phyto_event.csv"),
                            data.table = FALSE,
                            na.strings = "")
```

Some of the salinity zones (`salzone`) differ from the categories in [@LacouturePhytoplanktonindexbiotic2006]. Salinity zones are converted to one of four categories: "f" (freshwater), "m" (mesohaline), "o" (oligohaline), and "p" (polyhaline).
```{r}
events.df <- events.df %>% 
  mutate(sampledate = as.Date(sampledate),
         salzone = case_when(
           salzone %in% c("tf", "fe") ~ "f",
           salzone %in% c("m", "me") ~ "m",
           salzone %in% c("o", "oe") ~ "o",
           salzone %in% c("p", "pe") ~ "p",
           TRUE ~ as.character(NA)
         ))
```

Retain only unique information regarding the data provider (`source`), station name (`station`), sample date (`sampledate`), water layer (`layer`), and salinity zone (`salzone`).
```{r}
events.sub <- events.df %>% 
  select(source, station, sampledate, layer, salzone) %>% 
  distinct()
```

Import the salinity zone used by Jacqueline M. Johnson of the Chesapeake Bay Program. This will provide a check of salinity zone classification from 10 years ago to the salinity zone classification in CEDR.
```{r}
old.salzone <- readxl::read_excel(file.path(project.dir, "data/jackie_data/JMJ_PIBI_Salzone_Data.xlsx"),
                         sheet = "JMJ Salzone+Scores") %>% 
  clean_up() %>% 
  select(station, sample_date, ibi_salzone) %>% 
  mutate(sample_date = as.Date(sample_date)) %>% 
  rename(old_salzone = ibi_salzone,
         sampledate = sample_date) %>% 
  distinct()
```

Append the salinity zone defined by Jacqueline M. Johnson (`old_salzone`) to `wq.df`.
```{r}
events.sub <- inner_join(events.sub, old.salzone, by = c("station", "sampledate"))
```

Identify rows where the salinity zone specified by CEDR (`salzone`) differs from the salinity zone specified by Jacqueline M. Johnson (`old_salzone`).
```{r}
events.sub <- events.sub %>% 
  mutate(salzone_disagree = if_else(salzone != old_salzone, TRUE, FALSE)) %>% 
  filter(layer != "bp")

DT::datatable(events.sub)
```

Summarize and plot the differences in specified salinity zones.
```{r}
salzone.diff <- events.sub %>% 
  select(station, sampledate, source, salzone, old_salzone, salzone_disagree) %>% 
  unite(sal_group, salzone, old_salzone) %>% 
  distinct() %>% 
  filter(salzone_disagree == TRUE) %>% 
  group_by(sal_group) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  arrange(count) %>% 
  mutate(sal_group = factor(sal_group, levels = unique(sal_group))) %>% 
  ggplot(aes(sal_group, count, fill = count)) + 
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  geom_bar(stat = "identity") +
  xlab("Salinity Zone Disagreement") +
  ylab("Number of Disagreements")
salzone.diff
```

To be consitent with the data used to develope the PIBI, salinity zones specified by Jacqueline M. Johnson (`old_salzone`) were favored when salinity zones differed. To be more explicit, `salzone` is replaced by `old_salzone` when salinity zones differed.
```{r}
events.sub <- events.sub %>% 
  rename(cedr_salzone = salzone) %>% 
  mutate(salzone = if_else(salzone_disagree == TRUE, old_salzone, cedr_salzone))
```

Join `bay.df` and `events.sub` to combine phytoplankton count data with event data.
```{r}
bay.df <- left_join(bay.df, events.sub, by = c("source", "station",
                                               "sampledate", "layer")) %>% 
  mutate(unique_id = paste(unique_id, season, salzone, sep = "_"))
```