---
title: "R Notebook"
output: html_document
---
```{r echo=FALSE}
knitr::opts_chunk$set(eval=evaluate, cache=cache.me)
```

### Scoring Functions

The code below seems like a lot of duplication but rating thresholds for each index are very specific and sometimes even the same metric used in two different indices is not scored using the same logic. For example, in the spring  freshwater index total biomass (`total_phyto_biomass`) is scored as either a 1, 3, or 5, while in the spring mesohaline index total biomass (`total_phyto_biomass`) is scored as either a 1 or `NA`. Due to these discrepancies, I have chosen to create a separate function specific to each phytoplankton index.

Each of the functions follows the same logic. `metrics.long` is filtered to only represent a single season and salinity zone (`salzone`). Additionally, `metrics.long` is filtered to only include the metrics associated with the specified season and salinity zone. If any of the required metrics are missing from the data, then an error will be returned specifying that the function in question requires the following metric(s). Finally, the scoring thresholds found in an old Microsoft Access database created by Jacqueline M. Johnson of the Chesapeake Bay Program are applied to each of the metrics. I first tried to use the thresholds provided in Lacouture et al. [-@LacouturePhytoplanktonindexbiotic2006, pp. 602-203] but minor discrepancies, mostly rounding differences, resulted in more errors later in this document; therefore, Jacqueline M. Johnson's thresholds were favored. Originally, summer mesohaline and polyhaline indices included a picoplankton abundance metric. The picoplankton abundance metric has been dropped because picoplankton are no longer collected by the data providers. A warning will still appear if `pico.warning` is not set to `FALSE` indicating that their was no value for picoplankton abundance and therefore no score could be assigned.

Each of the scoring functions does not require a description because of the paragraph above, so only a subheading will be provided to break up the functions.

#### Spring Freshwater
```{r}
score_spring_f <- function(metrics.long) {
  metric.vec <- c("total_phyto_biomass_chla_ratio",
                  "surface_chla",
                  "cyanophyte_biomass",
                  "doc",
                  "pheophytin",
                  "total_phyto_biomass")
  #----------------------------------------------------------------------------
  metrics.sub <- metrics.long %>% 
    filter(season == "spring", 
           salzone == "f",
           metric %in% metric.vec)
  #----------------------------------------------------------------------------
  if (any(!metric.vec %in% unique(metrics.sub$metric))) {
    stop(paste("score_spring_f requires the following metric(s):",
               paste(metric.vec[!metric.vec %in% unique(metrics.sub$metric)], collapse = ", ")))
  }
  #----------------------------------------------------------------------------
  final.df <- metrics.sub %>% 
    mutate(score = case_when(
      # Jackie's code 39.28 and 41.72
      # Lacouture (2006) 39.3 and 41.7
      metric == "total_phyto_biomass_chla_ratio" & value < 39.28 ~ 1,
      metric == "total_phyto_biomass_chla_ratio" & value >= 39.28 & value <= 41.72 ~ 3,
      metric == "total_phyto_biomass_chla_ratio" & value > 41.72 ~ 5,
      metric == "surface_chla" & (value < 3.42 | value > 14.45) ~ 1,
      metric == "surface_chla" & ((value >= 3.42 & value <= 4.37) |(value >= 13.98 & value <= 14.45)) ~ 3,
      metric == "surface_chla" & value > 4.37 & value < 13.98 ~ 5,
      # Jackie's code 23.02
      # Lacouture (2006) 23
      metric == "cyanophyte_biomass" & value > 23.02 ~ 1,
      metric == "cyanophyte_biomass" & value <= 23.02 ~ as.numeric(NA),
      metric == "doc" & value > 2.40 ~ 1,
      metric == "doc" & value >= 2.19 & value <= 2.40 ~ 3,
      metric == "doc" & value < 2.19 ~ 5,
      metric == "pheophytin" & value > 2.50 ~ 1,
      metric == "pheophytin" & value >= 1.55 & value <= 2.50 ~ 3,
      metric == "pheophytin" & value < 1.55 ~ 5,
      # Jackie's code 172.99, 583.9, and 828.5
      # Lacouture (2006) 173, 584, and 829
      metric == "total_phyto_biomass"  & (value <= 172.99 | value > 828.5) ~ 1,
      metric == "total_phyto_biomass"  & value >= 583.9 & value <= 828.5 ~ 3,
      metric == "total_phyto_biomass"  & value > 172.99 & value < 583.9 ~ 5,
      TRUE ~ as.numeric(NA)))
  return(final.df)
}

```

#### Spring Oligohaline
```{r}
score_spring_o <- function(metrics.long) {
  metric.vec <- c("total_phyto_biomass_chla_ratio",
                  "surface_chla",
                  "doc",
                  "pheophytin",
                  "total_phyto_biomass")
  #----------------------------------------------------------------------------
  metrics.sub <- metrics.long %>% 
    filter(season == "spring", 
           salzone == "o",
           metric %in% metric.vec)
  #----------------------------------------------------------------------------
  if (any(!metric.vec %in% unique(metrics.sub$metric))) {
    stop(paste("score_spring_o requires the following metric(s):",
               paste(metric.vec[!metric.vec %in% unique(metrics.sub$metric)], collapse = ", ")))
  }
  #----------------------------------------------------------------------------
  final.df <- metrics.sub %>%
    mutate(score = case_when(
      metric == "total_phyto_biomass_chla_ratio" & (value < 18.8 | value > 48.6) ~ 1,
      metric == "total_phyto_biomass_chla_ratio" & value >= 18.8 & value <= 21.2 ~ 3,
      metric == "total_phyto_biomass_chla_ratio" & value > 21.2 & value < 48.6 ~ 5,
      metric == "surface_chla" & (value < 6.77 | value > 33.64) ~ 1,
      metric == "surface_chla" & ((value >= 20.93 & value <= 33.64) |(value >= 6.77 & value <= 8.82)) ~ 3,
      metric == "surface_chla" & value > 8.82 & value < 20.93 ~ 5,
      metric == "doc" & value > 3.27 ~ 1,
      metric == "doc" & value >= 2.69 & value <= 3.27 ~ 3,
      metric == "doc" & value < 2.69 ~ 5,
      metric == "pheophytin" & value > 2.68 ~ 1,
      metric == "pheophytin" & value >= 2.23 & value <= 2.68 ~ 3,
      metric == "pheophytin" & value < 2.23 ~ 5,
      # Jackie's code 133.67, 426.31, 131.37, and 685.81
      # Lacouture (2006) 134, 426, 131, and 686
      metric == "total_phyto_biomass"  & value > 133.67 & value < 426.31 ~ 5,
      metric == "total_phyto_biomass"  & ((value >= 131.37 & value <= 133.67) | (value >= 426.31 & value < 685.81)) ~ 3,
      metric == "total_phyto_biomass"  & any(value < 131.37 | value >= 685.81) ~ 1,
      TRUE ~ as.numeric(NA)))
  return(final.df)
}

```

#### Spring Mesohaline
```{r}
score_spring_m <- function(metrics.long) {
  metric.vec <- c("total_phyto_biomass_chla_ratio",
                  "surface_chla",
                  "diatom_biomass",
                  "dinoflagellate_biomass",
                  "doc",
                  "pheophytin",
                  "prorocentrum_minimum_abundance",
                  "total_phyto_biomass")
  #----------------------------------------------------------------------------
  metrics.sub <- metrics.long %>% 
    filter(season == "spring", 
           salzone == "m",
           metric %in% metric.vec)
  #----------------------------------------------------------------------------
  if (any(!metric.vec %in% unique(metrics.sub$metric))) {
    stop(paste("score_spring_m requires the following metric(s):",
               paste(metric.vec[!metric.vec %in% unique(metrics.sub$metric)],
                     collapse = ", ")))
  }
  #----------------------------------------------------------------------------
  final.df <- metrics.sub %>% 
    mutate(score = case_when(
      # Jackie's code 69.52 and 45.04
      # Lacouture (2006) 69.5 and 45
      metric == "total_phyto_biomass_chla_ratio" & value < 45.04 ~ 1,
      metric == "total_phyto_biomass_chla_ratio" & value >= 45.04 & value <= 69.52 ~ 3,
      metric == "total_phyto_biomass_chla_ratio" & value > 69.52 ~ 5,
      metric == "surface_chla" & (value < 2.60 | value > 8.00) ~ 1,
      metric == "surface_chla" & ((value >= 2.60 & value <= 2.90) |(value >= 6.17 & value <= 8.00)) ~ 3,
      metric == "surface_chla" & value > 2.90 & value < 6.17 ~ 5,
      # Jackie's code 275.4
      # Lacouture (2006) 275
      metric == "diatom_biomass" & (value < 149 | value >= 2513) ~ 1,
      metric == "diatom_biomass" & value >= 149 & value <= 275.4 ~ 3,
      metric == "diatom_biomass" & value > 275.4 & value < 2513 ~ 5,
      # Jackie's code 28.3, 156.9, and 268.2
      # Lacouture (2006) 157 and 268
      metric == "dinoflagellate_biomass" & (value < 28.3 | value > 268.2) ~ 1,
      metric == "dinoflagellate_biomass" & value >= 156.9 & value <= 268.2 ~ 3,
      metric == "dinoflagellate_biomass" & value > 28.3 & value < 156.9 ~ 5,
      metric == "doc" & value > 3.17 ~ 1,
      metric == "doc" & value >= 2.84 & value <= 3.17 ~ 3,
      metric == "doc" & value < 2.84 ~ 5,
      metric == "pheophytin" & value > 1.03 ~ 1,
      metric == "pheophytin" & value >= 1.00 & value <= 1.03 ~ 3,
      metric == "pheophytin" & value < 1.00 ~ 5,
      metric == "prorocentrum_minimum_abundance" & value > 1477600 ~ 1,
      metric == "prorocentrum_minimum_abundance" & value <= 1477600 ~ as.numeric(NA),
      metric == "total_phyto_biomass"  & value > 1150 ~ 1,
      metric == "total_phyto_biomass"  & value <= 1150 ~ as.numeric(NA),
      TRUE ~ as.numeric(NA)))
  return(final.df)
}

```

#### Spring Polyhaline
```{r}
score_spring_p <- function(metrics.long) {
  metric.vec <- c("total_phyto_biomass_chla_ratio",
                  "surface_chla",
                  "pct_cryptophyte",
                  "doc",
                  "pheophytin",
                  "prorocentrum_minimum_abundance",
                  "total_phyto_biomass")
  #----------------------------------------------------------------------------
  metrics.sub <- metrics.long %>% 
    filter(season == "spring", 
           salzone == "p",
           metric %in% metric.vec)
  #----------------------------------------------------------------------------
  if (any(!metric.vec %in% unique(metrics.sub$metric))) {
    stop(paste("score_spring_p requires the following metric(s):",
               paste(metric.vec[!metric.vec %in% unique(metrics.sub$metric)], collapse = ", ")))
  }
  #----------------------------------------------------------------------------
  final.df <- metrics.sub %>% 
    mutate(score = case_when(
      metric == "total_phyto_biomass_chla_ratio" & value < 71.0 ~ 1,
      metric == "total_phyto_biomass_chla_ratio" & value >= 71.0 & value <= 107.5 ~ 3,
      metric == "total_phyto_biomass_chla_ratio" & value > 107.5 ~ 5,
      metric == "surface_chla" & value > 4.00 ~ 1,
      metric == "surface_chla" & value >= 2.80 & value <= 4.00 ~ 3,
      metric == "surface_chla" & value < 2.80 ~ 5,
      # Jackie's code 4.93 and 7.06
      # Lacouture (2006) 4.9 and 7.1
      metric == "pct_cryptophyte" & value > 7.06 ~ 1,
      metric == "pct_cryptophyte" & value >= 4.93 & value <= 7.06 ~ 3,
      metric == "pct_cryptophyte" & value < 4.93 ~ 5,
      metric == "doc" & value > 2.61 ~ 1,
      metric == "doc" & value >= 2.50 & value <= 2.61 ~ 3,
      metric == "doc" & value < 2.50 ~ 5,
      metric == "pheophytin" & value > 0.90 ~ 1,
      metric == "pheophytin" & value >= 0.55 & value <= 0.90 ~ 3,
      metric == "pheophytin" & value < 0.55 ~ 5,
      metric == "prorocentrum_minimum_abundance" & value > 7488 ~ 1,
      metric == "prorocentrum_minimum_abundance" & value >= 672 & value <= 7488 ~ 3,
      metric == "prorocentrum_minimum_abundance" & value < 672 ~ 5,
      # Jackie's code 1061.7
      # Lacouture (2006) 1062
      metric == "total_phyto_biomass"  & value > 1061.7 ~ 1,
      metric == "total_phyto_biomass"  & value <= 1061.7 ~ as.numeric(NA),
      TRUE ~ as.numeric(NA)))
  return(final.df)
}

```

#### Summer Freshwater
```{r}
score_summer_f <- function(metrics.long) {
  metric.vec <- c("surface_chla",
                  "cyanophyte_biomass",
                  "diatom_biomass",
                  "doc", 
                  "microcystis_aeruginosa_abundance",
                  "pheophytin", 
                  "total_phyto_biomass")
  #----------------------------------------------------------------------------
  metrics.sub <- metrics.long %>% 
    filter(season == "summer", 
           salzone == "f",
           metric %in% metric.vec)
  #----------------------------------------------------------------------------
  if (any(!metric.vec %in% unique(metrics.sub$metric))) {
    stop(paste("score_summer_f requires the following metric(s):",
               paste(metric.vec[!metric.vec %in% unique(metrics.sub$metric)], collapse = ", ")))
  }
  #----------------------------------------------------------------------------
  final.df <- metrics.sub %>% 
    mutate(score = case_when(
      metric == "surface_chla" & value > 12.30 ~ 1,
      metric == "surface_chla" & ((value >= 12.00 & value <= 12.30) | value <= 5.4) ~ 3,
      metric == "surface_chla" & value > 5.40 & value < 12.00 ~ 5,
      # Jackie's code 38.87 and 67.4
      # Lacouture (2006) 39 and 67
      metric == "cyanophyte_biomass" & value > 67.4 ~ 1,
      metric == "cyanophyte_biomass" & value >= 38.87 & value <= 67.4 ~ 3,
      metric == "cyanophyte_biomass" & value < 38.87 ~ 5,
      # Jackie's code 122.1 and 192.6
      # Lacouture (2006) 122 and 193
      metric == "diatom_biomass" & value > 192.6 ~ 1,
      metric == "diatom_biomass" & value >= 122.1 & value <= 192.6 ~ 3,
      metric == "diatom_biomass" & value < 122.1 ~ 5,
      metric == "doc" & value > 3.18 ~ 1,
      metric == "doc" & value >= 2.67 & value <= 3.18 ~ 3,
      metric == "doc" & value < 2.67 ~ 5,
      metric == "microcystis_aeruginosa_abundance" & value > 262507 ~ 1,
      metric == "microcystis_aeruginosa_abundance" & value <= 262507 ~ as.numeric(NA),
      metric == "pheophytin" & value > 4.30 ~ 1,
      metric == "pheophytin" & value >= 2.40 & value <= 4.30 ~ 3,
      metric == "pheophytin" & value < 2.40 ~ 5,
      # Jackie's code 231.3 and 555.7
      # Lacouture (2006) 232 and 551
      metric == "total_phyto_biomass"  & value > 555.7 ~ 1,
      metric == "total_phyto_biomass"  & value < 231.3 ~ 3,
      metric == "total_phyto_biomass"  & value >= 231.3 & value <= 555.7 ~ 5,
      TRUE ~ as.numeric(NA)))
  return(final.df)
}
```

#### Summer Oligohaline
```{r}
score_summer_o <- function(metrics.long) {
  metric.vec <- c("surface_chla",
                  "cyanophyte_biomass",
                  "diatom_biomass",
                  "doc",
                  "pheophytin")
  #----------------------------------------------------------------------------
  metrics.sub <- metrics.long %>% 
    filter(season == "summer", 
           salzone == "o",
           metric %in% metric.vec)
  #----------------------------------------------------------------------------
  if (any(!metric.vec %in% unique(metrics.sub$metric))) {
    stop(paste("score_summer_o requires the following metric(s):",
               paste(metric.vec[!metric.vec %in% unique(metrics.sub$metric)], collapse = ", ")))
  }
  #----------------------------------------------------------------------------
  final.df <- metrics.sub %>% 
    mutate(score = case_when(
      metric == "surface_chla" & value >= 9.47 ~ 1,
      metric == "surface_chla" & value <= 4.20 ~ 3,
      metric == "surface_chla" & value > 4.20 & value < 9.47 ~ 5,
      # Jackie's code 1.79 and 26.55
      # Lacouture (2006) 1.8 and 27
      metric == "cyanophyte_biomass" & value >= 26.55 ~ 1,
      metric == "cyanophyte_biomass" & value < 1.79 ~ 3,
      metric == "cyanophyte_biomass" & value >= 1.79 & value < 26.55 ~ 5,
      # Jackie's code 44.14 and 126.59
      # Lacouture (2006) 44 and 127
      metric == "diatom_biomass" & value >= 126.59 ~ 1,
      metric == "diatom_biomass" & value <= 44.14 ~ 3,
      metric == "diatom_biomass" & value > 44.14 & value < 126.59 ~ 5,
      metric == "doc" & value > 4.00 ~ 1,
      metric == "doc" & value >= 3.15 & value <= 4.00 ~ 3,
      metric == "doc" & value < 3.15 ~ 5,
      metric == "pheophytin" & value > 2.81 ~ 1,
      metric == "pheophytin" & value >= 1.58 & value <= 2.81 ~ 3,
      metric == "pheophytin" & value < 1.58 ~ 5,
      TRUE ~ as.numeric(NA)))
  return(final.df)
}
```

#### Summer Mesohaline
```{r}
score_summer_m <- function(metrics.long, pico.warning) {
  metric.vec <- c("total_phyto_biomass_chla_ratio",
                  "surface_chla",
                  "dinoflagellate_biomass", 
                  "doc",
                  "pheophytin",
                  "total_phyto_biomass")
  pico.vec <- "picoplankton_abundance"
  #----------------------------------------------------------------------------
  metrics.sub <- metrics.long %>% 
    filter(season == "summer", 
           salzone == "m",
           metric %in% c(metric.vec, pico.vec))
  #----------------------------------------------------------------------------
  if (any(!metric.vec %in% unique(metrics.sub$metric))) {
    stop(paste("score_summer_m requires the following metric(s):",
               paste(metric.vec[!metric.vec %in% unique(metrics.sub$metric)], collapse = ", ")))
  }
  
  if (pico.warning == TRUE && any(!pico.vec %in% unique(metrics.sub$metric))) {
    warning("Warning: picoplankton_abundance missing from score_summer_m")
  }
  #----------------------------------------------------------------------------
  final.df <- metrics.sub %>% 
    mutate(score = case_when(
      metric == "total_phyto_biomass_chla_ratio" & value < 32.2 ~ 1,
      metric == "total_phyto_biomass_chla_ratio" & value >= 32.2 & value <= 36.9 ~ 3,
      metric == "total_phyto_biomass_chla_ratio" & value > 36.9 ~ 5,
      metric == "surface_chla" & value >= 9.74 ~ 1,
      metric == "surface_chla" & ((value >= 7.70 & value < 9.74) | value <= 4.00) ~ 3,
      metric == "surface_chla" & value > 4.00 & value < 7.70 ~ 5,
      # Jackie's code 200.92, 31.22, 55.98
      # Lacouture (2006) 201, 31, 56
      metric == "dinoflagellate_biomass" & (value <= 31.22 | value > 200.92) ~ 1,
      metric == "dinoflagellate_biomass" & value > 31.22 & value <= 55.98 ~ 3, 
      metric == "dinoflagellate_biomass" & value > 55.98 & value < 200.92 ~ 5,
      metric == "doc" & value > 3.35 ~ 1,
      metric == "doc" & value >= 2.99 & value <= 3.35 ~ 3,
      metric == "doc" & value < 2.99 ~ 5,
      metric == "pheophytin" & value > 1.60 ~ 1,
      metric == "pheophytin" & value >= 1.23 & value <= 1.60 ~ 3,
      metric == "pheophytin" & value < 1.23 ~ 5,
      # Jackie's code 598720000
      # Lacouture (2006) 598700000
      metric == "picoplankton_abundance" & value < 352000000 ~ 1,
      metric == "picoplankton_abundance" & value >= 352000000 & value <= 598720000 ~ 3,
      metric == "picoplankton_abundance" & value > 598720000 ~ 5,
      metric == "total_phyto_biomass"  & value > 660 ~ 1,
      metric == "total_phyto_biomass"  & value <= 660 ~ as.numeric(NA),
      TRUE ~ as.numeric(NA)))
  return(final.df)
}
```

#### Summer Polyhaline
```{r}
score_summer_p <- function(metrics.long, pico.warning) {
  
  metric.vec <- c("total_phyto_biomass_chla_ratio",
                  "surface_chla",
                  "pct_cryptophyte",
                  "diatom_biomass",
                  "dinoflagellate_biomass",
                  "doc",
                  "pheophytin",
                  "total_phyto_biomass")
  pico.vec <- "picoplankton_abundance"
  #----------------------------------------------------------------------------
  metrics.sub <- metrics.long %>% 
    filter(season == "summer", 
           salzone == "p",
           metric %in% c(metric.vec, pico.vec))
  #----------------------------------------------------------------------------
  if (any(!metric.vec %in% unique(metrics.sub$metric))) {
    stop(paste("score_summer_p requires the following metric(s):",
               paste(metric.vec[!metric.vec %in% unique(metrics.sub$metric)], collapse = ", ")))
  }
  
  if (pico.warning == TRUE && any(!pico.vec %in% unique(metrics.sub$metric))) {
    warning("Warning: picoplankton_abundance missing from score_summer_p")
  }
  #----------------------------------------------------------------------------
  final.df <- metrics.sub %>% 
    mutate(score = case_when(
      metric == "total_phyto_biomass_chla_ratio" & value < 37.7 ~ 1,
      metric == "total_phyto_biomass_chla_ratio" & value >= 37.7 & value <= 74.5 ~ 3,
      metric == "total_phyto_biomass_chla_ratio" & value > 74.5 ~ 5,
      metric == "surface_chla" & value > 5.33 ~ 1,
      metric == "surface_chla" & value >= 4.52 & value <= 5.33 ~ 3,
      metric == "surface_chla" & value < 4.52 ~ 5,
      metric == "pct_cryptophyte" & value > 6.5 ~ 1,
      metric == "pct_cryptophyte" & value >= 3.9 & value <= 6.5 ~ 3,
      metric == "pct_cryptophyte" & value < 3.9 ~ 5,
      metric == "diatom_biomass" & (value >= 799 | value < 137) ~ 1,
      metric == "diatom_biomass" & value >= 137 & value <= 181 ~ 3,
      metric == "diatom_biomass" & value > 181 & value < 799 ~ 5,
      metric == "dinoflagellate_biomass" & (value < 23 | value >= 544) ~ 1,
      metric == "dinoflagellate_biomass" & value >= 23 & value <= 37 ~ 3, 
      metric == "dinoflagellate_biomass" & value > 37 & value < 554 ~ 5,
      metric == "doc" & value > 2.80 ~ 1,
      metric == "doc" & value >= 2.58 & value <= 2.80 ~ 3,
      metric == "doc" & value < 2.58 ~ 5,
      metric == "pheophytin" & value > 1.50 ~ 1,
      metric == "pheophytin" & value >= 0.93 & value <= 1.50 ~ 3,
      metric == "pheophytin" & value < 0.93 ~ 5,
      metric == "picoplankton_abundance" & value < 208600000 ~ 1,
      metric == "picoplankton_abundance" & value >= 208600000 & value <= 269500000 ~ 3,
      metric == "picoplankton_abundance" & value > 269500000 ~ 5,
      # In Jackie's code 718
      # Lacouture (2006) 711
      metric == "total_phyto_biomass"  & (value < 181 | value > 831) ~ 1,
      metric == "total_phyto_biomass"  & ((value >= 181 & value <= 207) | (value >= 718 & value <= 831)) ~ 3,
      metric == "total_phyto_biomass"  & value > 207 & value < 718 ~ 5,
      TRUE ~ as.numeric(NA)))
  return(final.df)
}
```

#### Scoring Wrapper Function

The following function wraps all of the functions above into a single function to make it simple to calculate scores. The `unique_id` column is separated to represent  station name (`station`), water layer (`layer`), sample number (`samplenumber`), sample date (`sampledate`), season (`season`), and salinity zone (`salzone`). Many of these fields are necessary for the individual scoring functions (e.g. `score_spring_f()` and `score_summer_p()`). The individual scoring functions are designed to score only the data applicable to a specific season and salinity zone; therefore, the output of each of the functions can simply be appended together.
```{r}
score_phyto <- function(metrics.long, pico.warning) {
  #----------------------------------------------------------------------------
  spring.f <- score_spring_f(metrics.long)
  spring.o <- score_spring_o(metrics.long)
  spring.m <- score_spring_m(metrics.long)
  spring.p <- score_spring_p(metrics.long)
  #----------------------------------------------------------------------------
  summer.f <- score_summer_f(metrics.long)
  summer.o <- score_summer_o(metrics.long)
  summer.m <- score_summer_m(metrics.long, pico.warning)
  summer.p <- score_summer_p(metrics.long, pico.warning)
  #----------------------------------------------------------------------------
  final.df <- bind_rows(spring.f, spring.o, spring.m, spring.p,
                        summer.f, summer.o, summer.m, summer.p) %>% 
    select(-layer)
  #----------------------------------------------------------------------------
  return(final.df)
}
```

### Apply the Scoring Function

`score_phyto()` is used to score the metric values in `metrics.long`.
```{r}
scores.df <- metrics.long %>% 
  separate(unique_id, c("station", "layer", "samplenumber",
                        "sampledate", "season", "salzone"), sep = "_",
           remove = FALSE) %>% 
  rename(date = sampledate) %>% 
  score_phyto(pico.warning = FALSE) %>% 
  select(-samplenumber)
```

The IBI score (`ibi_score`) is calculated by finding the mean score for each unique event (`unique_id`), excluding `NA`s. Additionally, metrics with a value that was not `NA` are counted (`present_count`). Unique events (`unique_id`) with fewer than 4 non-`NA` values are excluded from the data frame.
```{r}
scores.df <- scores.df %>% 
  group_by(unique_id) %>% 
  mutate(ibi_score = mean(score, na.rm = TRUE),
         metric_count = n(),
         present_count = sum(!is.na(score)),
         na_count = sum(is.na(score))) %>% 
  ungroup() %>% 
  filter(present_count >= 4)
```

## Ratings

All of the indices were rated using the same rating scheme [@LacouturePhytoplanktonindexbiotic2006, p. 611].

* **poor:** score <= 2
* **fair-poor:** 2 < score < 2.67
* **fair:** 2.67 <= score < 3.33
* **fair-good:** 3.33 <= score <= 4
* **good:** score > 4
```{r}
rate_phyto <- function(scores.df) {
  final.df <- scores.df %>% 
    mutate(
      ibi_score = round(ibi_score, 2),
      rating = case_when(
        ibi_score < 2 ~ "poor",
        ibi_score >= 2 & ibi_score < 2.67 ~ "fair_poor",
        ibi_score >= 2.67 & ibi_score < 3.33 ~ "fair",
        ibi_score >= 3.33 & ibi_score < 4 ~ "fair_good",
        ibi_score >= 4 ~ "good",
        TRUE ~ "ERROR"
      ))
  return(final.df)
}
```

Apply `rate_phyto()` to `scores.df`.
```{r}
ratings.df <- rate_phyto(scores.df) %>% 
  mutate(date = as.Date(date))
```


